Composition-climate paper

### 2016-08-22 Composition_Ecology.Rmd
Changes up to the end of the Methods.  Added reference to Mann *et al*.

Estimates of settlement-era forest composition and tree distribution based on survey data from the Public Land Survey System [PLS: @white1983history;@schulte2001original;@gorng2016preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@gorng2016preset]. We briefly review the methods here. PLS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history]. At each survey, point surveyors noted the closest two to four trees and recorded their common name, distances to the survey point, diameters at breast height and azimuth to the tree. Goring et al. [-@gorng2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@gorng2016preset]. Using the count data produced by Goring et al. [-@gorng2016preset], we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (`r tab_nums("genera_table", display = "cite")`). Common names for certain taxa, such as "Ironwood" can not be clearly resolved to a single tree genera, as such we combine *Ostrya* and *Carpinus* (ironwood), and *Thuja* and *Juniperus* (cedar). Here presence is defined as at least one recorded tree in any 8x8 km grid cell.

Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which began in the 1930s and provide detailed inventories of contemporary tree distributions [@iverson2013tree;@zhu2012failure]. FIA datasets are publicly available surveys of forest composition across the United States [@woudenberg2010forest].  FIA uses a nationally standardized sampling procedure, for all land with greater than 10% forest cover.  Each Phase 2 and 3 FIA plot  consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@gorng2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here the primary tool for maximizing comparability is via aggregation, both by spatially aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  For example,  the FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data. This leads to higher cell-to-cell variability in the FIA data, but these are not expected to affect the regional-scale analyses here, particularly because we are investigating presence-absence patterns. Additionally, FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing the spatial extent of many tree taxa to be lower in the FIA than in the PLS datasets (`r tab_nums("genera_table", display="cite")`). Hence, scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA. However the practice of using FIA data as a surrogate for tree presence-absence data is widespread [@iverson2013tree;@iverson1998predicting], so this complication is a general one faced by forest ecologists and global change ecologists.  In prior comparative analyses of the PLS and FIA data, Goring et al. [-@gorng2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.

@@

Historical and modern climate data are derived from monthly PRISM data for 1895 to 2014 CE [@prism2004]. The PRISM climate variables are available at an 800m resolution and have been interpolatedfrom climate stationlocations using elevation, aspectand other data.  Estimates for July temperature ($T_{jul}$), January temperature ($T_{jan}$), the annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were obtained from PRISM and resampled to the 8 x 8km grid used for PLS and FIA data using bilinear interpolation. Two periods were selected for the calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.  There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years.  Few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period, although recent methodological advances [@tiptoninreview] may provide estimates from oberver datasets as far back as 1825.  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we check the instrumental record against tree-ring-based paleoclimate reconstructions from the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).

@@

Distributions of tree genera presences along climate gradients are described as probability density functions for both the pre-settlement and modern eras.  Differences in distributions are quantified using Hellinger distances [@matthes2016benchmarking] to compare the probability density functions among the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate and the subscripts M and H represent modern and historical conditions, respectively). The total shift ($d_{tot}$) between presettlement conditions and modern conditions is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ densities (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). The effect of historical climate change on shifting climate distributios of tree genera is determined by calculating the Hellinger distance ($d_{C}$) between PLS and FIA tree distributions when both are overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; `r fig_nums("fourPanel", display = "cite")`a; left column). We develop an attribution index, ($\delta_{d}=d_{c} - d_{l}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.

To assess whether the effects of historical land use and climatic changes on a genera's climatic distribution is compounding, counteracting or confounding we perform a set of t-tests, using the conservative Bonferroni correction to account for multiple tests.  For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (land-use change) distributions are significantly different.   If both changes are significant and of the same sign, then the shift is defined as a compounding change.  If both are significant and are in opposite directions then the change is defined as counteracting.  In the case that either or both changes are not significant then the change is defined as confounding.

Analysis in this paper is undertaken using R [@Team2014] and R Markdown [@rmarkdown2016].  All code and raw data is provided in supplemental material and hosted on GitHub (http://github.com/SimonGoring/Composition_Climate).  Analysis makes use of the R packages colorspace, captioner, raster, ggplot2, mgcv, reshape2, plyr, gridExtra, rgdal, RColorBrewer, analogue, and maptools.  Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape-level features.

@@

*The Results section should present the experiments that support the conclusions to be drawn later in the Discussion. The Results section should conform to a high standard of rigour. Extended lines of inference, arguments or speculations should not be placed in the Results.*

>>>>

Estimates of settlement-era forest composition and tree distribution based on survey data from the Public Land Survey System [PLS: @white1983history;@schulte2001original;@gorng2016preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@gorng2016preset]. We briefly review the methods here. PLS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history]. At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuths to each tree. Goring et al. [-@gorng2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@gorng2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (`r tab_nums("genera_table", display = "cite")`). A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus* to Ironwood, and *Thuja* and *Juniperus* to Cedar. 'Here 'Presence' is defined as at least one recorded tree in any 8x8 km grid cell.

Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which began in the 1930s and provides detailed inventories of contemporary tree distributions [@iverson2013tree;@zhu2012failure], for all regions with >10% forest cover [@woudenberg2010forest].  FIA uses a nationally standardized sampling procedure, that has evolved over time.  In the Phase 2/3 system, beginning in 1998, each plot consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@gorng2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  For example, the FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data. This leads to higher cell-to-cell variability in the FIA data, but these are not expected to affect the regional-scale analyses here, particularly because we are investigating presence-absence patterns. Additionally, FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing the spatial extent of many tree taxa to be lower in the FIA than in the PLS datasets (`r tab_nums("genera_table", display="cite")`). Hence, scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA. However the practice of using FIA data as a surrogate for tree presence-absence data is widespread [@iverson2013tree;@iverson1998predicting], so this complication is a general one faced by forest ecologists and global change ecologists.  In prior comparative analyses of the PLS and FIA data, Goring et al. [-@gorng2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.

@@

Historical and modern climate data are derived from monthly PRISM data for 1895 to 2014 CE [@prism2004]. The PRISM climate variables are available at an 800m resolution and have been interpolatedfrom climate stationlocations using elevation, aspectand other data.  Estimates for July temperature ($T_{jul}$), January temperature ($T_{jan}$), the annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were obtained from PRISM and resampled to the 8 x 8km grid used for PLS and FIA data using bilinear interpolation. Two periods were selected for the calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.  There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years.  This temporal mismatch is unavoidable; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against tree-ring-based paleoclimate reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).

@@

Distributions of tree genera presences in climate space are described as univariate density functions (PDFs) for both the pre-settlement and modern eras.  Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] to compare PDFs among the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate, and M and H represent modern and historical conditions). The total shift ($d_{tot}$) between pre-settlement and modern conditions is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ densities (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). The effect of historical climate change on shifting climate distributios of tree genera is determined by calculating the Hellinger distance ($d_{C}$) between PLS and FIA tree distributions when both are overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; `r fig_nums("fourPanel", display = "cite")`a; left column). Conversely, the effect of historical vegetation change on tree-climate distributions is determined by calculating the Hellinger distance ($d_{l}$) betweem the PLSS tree distributions when overlaid on historical vs. modern climates ($V_{H}C_{H}$ versus $V_{H}C_{M}$; Figure 2a; top row).  We develop an attribution index, ($\delta_{d}=d_{c} - d_{l}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.

To assess whether the effects of historical land use and climatic changes on a genera's climatic distribution is compounding, counteracting or confounding we perform a set of t-tests, using the conservative Bonferroni correction to account for multiple tests.  For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to land-use change) distributions are significantly different.   If both changes are significant and of the same sign, then the shift is defined as compounding.  If both are significant and are in opposite directions then the shift is counteracting.  In the case that either or both changes are insignificant, then the change is defined as confounding.

All analysis makes use of R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`. All code is hosted on GitHub (http://github.com/PalEON-Project/Composition_Climate). Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape-level features.

@@
